Clause Size Reduction with \alluip Learning

### 2020-05-07 paper.tex
change remaining i-uip to all-uip
fb camera ready revisions

\newcommand{\fs}[1]{\fontsize{#1}{#1}\selectfont}
\newcommand{\fss}[2]{\fontsize{#1}{#2}\selectfont}
\newcommand{\sat}{SAT\xspace}
\newcommand{\code}[1]{\text{#1}}
\newcommand{\assertionTrail}{\trail}
\newcommand{\LBD}{\text{LBD}\xspace}
\newcommand{\gap}{\text{Gap}}
\newcommand{\tgap}{t_{\mathit{gap}}}
\newcommand{\ntries}{n_{\mathit{tries}}}
\newcommand{\nsuc}{n_{\mathit{succ}}}
\newcommand{\allUip}{\textit{stable-allUIP}}
\newcommand{\tryuiplevel}{\textit{try-uip-level}\xspace}
\newcommand{\allUipPure}{\textit{pure-allUIP}\xspace}
\newcommand{\allUipMin}{\textit{min-allUIP}\xspace}
\newcommand{\allUipAct}{\textit{allUIP-Active}}
\newcommand{\allUipIn}{\textit{allUIP-Inclusive}}
\newcommand{\allUipEx}{\textit{allUIP-Exclusive}}
\newcommand{\MapleBase}{\textit{MapleCOMSPS\_LRB}}
\newcommand{\MapleSeven}{\textit{MapleLCMDist}}
\newcommand{\MapleNine}{\textit{MapleLCMDiscChronoBT-DL-v3} }
\newcommand{\MapleEight}{\textit{MapleLCMDiscChronoBT}}
\newcommand{\MapleNineShort}{\textit{MapleCB-DL} }
\newcommand{\expSAT}{\textit{expMaple\_CM\_GCBumpOnlyLRB} }
\newcommand{\expSATShort}{\textit{expMaple} }
\newcommand{\MapleIUIPPure}{\text{Maple-\allUipPure}}
\newcommand{\MapleIUIMin}{\text{Maple-\allUipMin}}
\newcommand{\MapleEightShort}{\textit{MapleCB}}
\newcommand{\Set}[2]{\{\,#1\mid#2\,\}}
\newcommand{\Array}[2]{[\,#1\mid#2\,]}
\newcommand{\target}{\textbf{Succ_t}}
\newcommand{\trail}{\ensuremath{\mathcal{T}}}
\newcommand{\trailIdx}[1]{\ensuremath{\iota(#1)}}
\newcommand{\dlevel}[1]{\ensuremath{\mathit{decLvl}(#1)}}
\newcommand{\dlevels}{\ensuremath{\mathit{decLvls}}}
\newcommand{\var}{\text{var}}
\newcommand{\true}{\textsc{true}\xspace}
\newcommand{\false}{\textsc{false}\xspace}
\newcommand{\reason}[1]{\ensuremath{\mathit{reason}(#1)}}
\newcommand{\resolve}{\bowtie}
\renewcommand{\implies}{\rightarrow}
\newcommand{\ctry}{C_{\mathit{try}}}
\newcommand{\type}{\textit{type}}
\newcommand{\unmarked}{\textbf{unmarked}}
\newcommand{\mrk}{\textbf{mark}}
\setlength{\theorempreskipamount}{3pt}
\setlength{\theorempostskipamount}{3pt}

@@

\title{Clause Size Reduction with all-UIP Learning}
\author{Nick Feng \and Fahiem Bacchus}
\authorrunning{N. Feng and F. Bacchus}
\institute{Department of Computer Science, University of Toronto, Canada, 
  \email{\{fengnick,fbacchus\}@cs.toronto.edu}}

\begin{document}
\maketitle              % typeset the header of the contribution
% 
\begin{abstract}
    Almost all CDCL \sat solvers use the 1-UIP 
    clause learning scheme for learning new clauses from conflicts, 
    and our current understanding
    of \sat solving provides good reasons for using that scheme. In
    particular, the 1-UIP scheme yields asserting clauses, and these
    asserting clauses have minimum LBD among all possible asserting
    clauses. As a result of these advantages, other clause learning
    schemes, like $i$-UIP and all-UIP, that were proposed in early
    work are not used in modern solvers. In this paper, we propose a
    new technique for exploiting the all-UIP clause learning scheme. 
    Our technique is to employ all-UIP learning under the
    constraint that the learnt clause's LBD does not increase (over the
    minimum established by the 1-UIP clause). Our method can learn
    clauses that are significantly smaller than the 1-UIP clause while
    preserving the minimum LBD. Unlike previous clause minimization
    methods, our technique is not limited to learning a sub-clause of
    the 1-UIP clause. It can instead learn a completely different
    clause. We show empirically 
    that our method can improve the performance of state of the art solvers.
\end{abstract}

\section{Introduction}
Clause learning is an essential technique in \sat solvers. There is
good evidence to indicate that it is, in fact, the most important
technique used in modern \sat solvers
\cite{DBLP:conf/sat/KatebiSS11}. In early \sat research a number of
different clause learning techniques were proposed
\cite{DBLP:conf/iccad/ZhangMMM01,DBLP:conf/iccad/SilvaS96,DBLP:journals/tc/Marques-SilvaS99,DBLP:conf/aaai/BayardoS97}.
However, following the revolutionary performance improvements achieved
by the Chaff \sat solver, the field has converged on using the 1-UIP
(first Unique Implication Point) scheme
\cite{DBLP:conf/iccad/ZhangMMM01} employed in Chaff
\cite{DBLP:conf/dac/MoskewiczMZZM01} (as well as other techniques
pioneered in the Chaff solver).\footnote{The idea of UIP clauses was
  first mentioned in \cite{DBLP:journals/tc/Marques-SilvaS99}, and
  1-UIP clauses along with other UIP clauses were learnt and used in
  the earlier GRASP \sat solver.} Since then almost all \sat solvers
have employed the 1-UIP clause learning scheme, along with clause
minimization \cite{DBLP:conf/sat/SorenssonB09}, as their primary
method for learning new clauses.

However, other clause learning schemes can be used in SAT solvers
without changes to the main data structures. Furthermore, advances in
our understanding allow us to better understand the potential
advantages and disadvantages of these alternate schemes. In this paper
we reexamine these previously proposed schemes with a focus on the
schemes described in \cite{DBLP:conf/iccad/ZhangMMM01}. Improved
understanding of \sat solvers, obtained from the last decade of
research, allows us to see that in their original form these
alternative clause learning schemes suffer significant disadvantages
over 1-UIP clause learning.

One of the previously proposed schemes was the all-UIP scheme
\cite{DBLP:conf/iccad/ZhangMMM01}. In this paper we propose a new way
to exploit the main ideas of this scheme that avoids its main
disadvantages. In particular, we propose to use a all-UIP like clause
learning scheme to generate smaller learnt clauses which retain the
good properties of standard 1-UIP clauses. Our method is related to,
but not the same as, various clause mininimization methods that try to
remove redundant literals from the 1-UIP clause yielding a clause that
is a subset of the 1-UIP clause, e.g.,
\cite{DBLP:conf/sat/SorenssonB09,DBLP:conf/ijcai/LuoLXML17,DBLP:conf/sat/WieringaH13}.
Our method is orthogonal to clause minimization. In particular, our
approach can learn a clause that is not a subset of the 1-UIP clause
but which still serves all of the same purposes as the 1-UIP clause.
Various minimization techniques can be applied on top of our method to
remove redundant literals from the clauses we have learnt.

We present various versions of our method and show that these variants
are often capable of learning shorter clauses than the 1-UIP scheme,
and that this can lead to useful performance gains in state of the art
\sat solvers.

@@

Two types of true literals appear on the trail: \emph{decision
  literals} that have been assumed to be true by the solver, and
\emph{unit propagated literals} that are forced to be true because
they are the sole remaining unfalsified literal of a clause. Each
literal $\ell\in\trail$ has a decision level $\dlevel{\ell}$ which is
equal to the number of decision literals appearing before $\ell$ on
the trail plus one. Hence, $\dlevel{d}=1$ for the first decision
literal $d\in\trail$.  The set of literals on $\trail$ that have the
same decision level forms a contiguous subsequence of $\trail$ that
starts with a decision literal $d_i$ and ends just before the next
decision literal $d_{i+1}$. If $\dlevel{d_i} = i$ we call this
subsequence of $\trail$ the \textit{$i$-th decision level}.

@@

In most \sat solvers, clause learning is initiated as soon as a clause
is falsified by $\trail$. In this paper we will be concerned with the
subsequent clause learning process which uses $\trail$ to derive a new
clause. In some implementations the trail might be altered during
clause learning. Here, however, we will assume that $\trail$
\emph{remains intact during clause learning} and is only changed after
the new clause is learnt. After the new clause is learnt $\trail$ will
be changed by backtracking.

Say that $\trail$ falsifies a clause $C_I$, and that the last decision
literal $d_k$ in $\trail$ has decision level $k$. Consider
$\trail_{k-1}$ the prefix of $\trail$ above the last decision level,
i.e., the sequence of literals
$\trail[0]$---$\trail[\trailIdx{d_k}-1]$. We will assume that
$\trail_{k-1}$ is \textbf{propagation complete}, although the full
trail $\trail$ might not be. This means that (a) no clause was
falsified by $\trail_{k-1}$. And (b) if $C_u$ is a clause containing
the literal $x$ and all literals in $C_u$ except for $x$ are falsified
by $\trail_{k-1}$, then $x\in \trail_{k-1}$ and
$\dlevel{x} \leq\max\{\dlevel{y} | y\in C_u \land y\neq x\}$. This
means that if $x$ appears in a clause made unit it must have been
added to the trail, and added at or before decision level the clause
became unit. Note that more than one clause might be made unit by
$\trail$ forcing $x$, or $x$ might be set as a decision before being
forced. This condition ensures that $x$ appears in $\trail$ at or
before the first decision level it is forced by any clause.

Any clause falsified by $\trail$ is called a \textbf{conflict}. When a
conflict is found, the final level of the trail, $k$, need not be
propagation complete as the solver typically stops propagation as soon
as it finds a conflict. This means that (a) other clauses might be
falsified by $\trail$ besides the conflict found, and (b) other
literals might be unit implied by $\trail$ but not added to $\trail$.

@@

\begin{Obs}
    If the unordered trail resolution $U$ yields the conflict clause
    $C_L$ from an initial conflict $C_I$, then there exists an ordered
    trail resolution $O$ that yields a conflict clause $C'_L$ such
    that $C'_L\subseteq C_L$.
\end{Obs}
\noindent
\textbf{Proof.} Let $U$ be the sequence of clauses $C_I = C_0$, $C_1$,
$\ldots$, $C_m= C_L$ obtained by resolving on the sequence of
variables $v_1$, $\ldots$, $v_m$ whose corresponding literals on
$\trail$ are $l_1$, $\ldots$, $l_m$. Reordering these resolution steps
so that the variables are resolved in order of decreasing trail index
and removing duplicates yields an ordered trail resolution $O$ with
the desired properties. Since no reason clause contains literals with
higher trail indices, $O$ must be a valid trail resolution if $U$
was, and furthermore $O$ yields the clause
$C'_L = \bigcup_{i=1}^m \reason{l_i} - \{l_1, \lnot l_1, \ldots, l_m,
\lnot l_m\}$. Since $U$ resolves on the same variables (in a different
order) using the same reason clauses we must have $C'_L\subseteq
C_L$. It can, however, be the case that $C'_L$ is proper subset of
$C_L$: if $l_i$ is resolved away it might be reintroduced when
resolving on $l_{i+1}$ if $\trailIdx{l_{i+1}}  > \trailIdx{l_i}$.
\whitebox

The relevance of trail resolutions is that all proposed clause
learning schemes we are aware of use trail resolutions to produce
learnt clauses. Furthermore, the commonly used technique for clause
minimization \cite{DBLP:conf/sat/SorenssonB09} is also equivalent to a
trail resolution that yields the minimized clause from the
un-minimized clause. Interestingly, it is standard in \sat solver
implementations to perform resolution going backwards along the
trail. That is, these implementations are typically using ordered
trail resolutions. The above observation shows that this in fact the
right way to do this.

@@

\subsection{Some Alternate Clause Learning Schemes}
A number of different clause learning schemes for generating a new
learnt clause from the initial conflict have been presented in prior
work, e.g.,
\cite{DBLP:conf/iccad/ZhangMMM01,DBLP:conf/iccad/SilvaS96,DBLP:journals/tc/Marques-SilvaS99,DBLP:conf/aaai/BayardoS97}.
Figure~\ref{fig:cl_schemes} gives a specification of some of these
methods: (a) the all-decision scheme which resolves away all implied
literals leaving a learnt clause over only decision literals; (c) the
1-UIP scheme which resolves away literals from the deepest decision
level leaving a learnt clause with a single literal at the deepest
level; (d) the all-UIP scheme which resolves away literals from each
decision level leaving a learnt clause with a single literal at each
decision level; and (e) the $i$-UIP scheme which resolves away
literals from the $i$ deepest decision levels leaving a learnt clause
with a single literal at its $i$ deepest decision levels. It should be
noted that when resolving away literals at decision level $i$, new
literals at decision levels less than $i$ might be introduced into the
clause. Hence, it is important in the $i$-UIP and all-UIP schemes to
use ordered trail resolutions.

Both the all-decision and all-UIP schemes yield a clause with only one
literal at each decision level, and the all-UIP clause will be no
larger that the all-decision clause. Furthermore, it is known
\cite{DBLP:journals/tc/Marques-SilvaS99} that once we reduce the
number of literals at a decision level $d$ to one, we could continue
performing resolutions and later achieve a different single literal at
the level $d$. In particular, a decision level might contain more than
one unique implication point. The algorithms given in
Figure~\ref{fig:cl_schemes} stop at the first UIP of a level, except
for the all-decision schemes with stops at the last UIP of each level.

\begin{figure}[t!]
{\fss{8pt}{10pt}
\begin{tabular}[t]{|l|c|l|}
  \cline{1-1}\cline{3-3}
  \begin{minipage}[t]{0.48\textwidth}
    \textbf{(a) All Decision Clause\rule{0pt}{1.1\topskip}}
    \begin{algorithmic}[0]
        \State \hspace*{-1em}\textbf{all\_decision}($C_I$)
        \State $C\gets C_I$
        \While{$\{l\,|\, l\in C \land\reason{l} \neq\varnothing\} \neq \emptyset$}
        \State \begin{tabular}{ll}
                 $\ell \gets \mbox{}$ & literal with highest trail index in\\
                                      & $\{l\,|\, l\in C \land\reason{l} \neq\varnothing\}$
               \end{tabular}
        \State $C \gets C\resolve \reason{\lnot \ell}$
        \EndWhile
        \State \textbf{return} $C$
    \end{algorithmic}
  \end{minipage}
  & &
  \begin{minipage}[t]{0.475\textwidth}
      \textbf{(b) Make level $i$ contain a single literal\rule{0pt}{1.1\topskip}}
      \begin{algorithmic}[0]
          \State \hspace*{-1em} \textbf{UIP\_level}($C$, $i$)
          \While{$\left|\left\{\ell \left| \begin{array}{l}\ell \in C\\
                                           \mbox{}\land \reason{l} \neq \varnothing \\
                                           \mbox{}\land \dlevel{l} = i\end{array}\right.\right\}\right| > 1$}
          \State \begin{tabular}{ll}
                   $l \gets \mbox{}$ & literal with highest trail index in\\
                                        & $\left\{\ell \left| \begin{array}{l}\ell \in C
                                          \land \reason{l} \neq \varnothing \\
                                          \mbox{}\land \dlevel{l} = i\end{array}\right.\right\}$
                 \end{tabular}
          \State $C \gets C\resolve \reason{\lnot l}$
          \EndWhile
          \State \textbf{return} $C$\strut
      \end{algorithmic}
  \end{minipage}
  \\\cline{1-1}\cline{3-3}
   \multicolumn{1}{c}{\vspace*{-7pt}}\\\cline{1-1}\cline{3-3}
  \begin{minipage}[t]{0.475\textwidth}
    \begin{minipage}{\textwidth}
        \textbf{(c) First UIP Clause \rule{0pt}{1.1\topskip}}
        \begin{algorithmic}[0]
            \State \hspace*{-1em}\textbf{1-UIP}($C_I$)
            \State $i\gets \max\{\dlevel{l}\, |\, l \in C_I\}$
            \State \textbf{return} \textbf{UIP\_level}($C_I$, i)
        \end{algorithmic}
        \vspace*{5pt}
    \end{minipage}\\[5pt]
    \hspace*{-1.8pt}\rule{1.03\textwidth}{.4pt}\\[5pt]
    \begin{minipage}{\textwidth}
        \textbf{(d) All UIP Clause \rule{0pt}{1.1\topskip}}
        \begin{algorithmic}[0]
          \State \hspace*{-1em}\textbf{all-UIP}($C_I$, $i$)
          \State $i = \bigl|\{\dlevel{l} \,|\, l \in \trail\}\bigl|$
          \State \Comment{$i$ is large enough to ensure all levels are UIP}
          \State \textbf{return} \textbf{$i$-UIP}(C, i)

@@

\subsection{Asserting Clauses and LBD---Reasons to Prefer 1-UIP
  Clauses}
An \textbf{asserting clause} \cite{DBLP:journals/ai/PipatsrisawatD11}
is a conflict clause $C_L$ that has exactly one literal $\ell$ at its
deepest level, i.e.,
$\forall x\in C_L. \dlevel{x} \leq \dlevel{\ell} \land (\dlevel{x} =
\dlevel{\ell} \implies x = \ell$). All of the clause learning schemes in
Figure~\ref{fig:cl_schemes} produced learnt clauses that are asserting.

>>>>

\newcommand{\fs}[1]{\fontsize{#1}{#1}\selectfont}
\newcommand{\fss}[2]{\fontsize{#1}{#2}\selectfont}
\newcommand{\sat}{SAT\xspace}
\newcommand{\code}[1]{\text{#1}}
\newcommand{\assertionTrail}{\trail}
\newcommand{\oneuip}{1-UIP\xspace}
\newcommand{\alluip}{all-UIP\xspace}
\newcommand{\LBD}{\text{LBD}\xspace}
\newcommand{\gap}{\text{gap}}
\newcommand{\tgap}{t_{\mathit{gap}}}
\newcommand{\ntries}{n_{\mathit{tries}}}
\newcommand{\nsuc}{n_{\mathit{succ}}}
\newcommand{\stablealluip}{\textit{stable-alluip}}
\newcommand{\tryuiplevel}{\textit{try-uip-level}\xspace}
\newcommand{\allUipPure}{\textit{pure-alluip}\xspace}
\newcommand{\allUipMin}{\textit{min-alluip}\xspace}
\newcommand{\allUipAct}{\textit{alluip-active}}
\newcommand{\allUipIn}{\textit{alluip-inclusive}}
\newcommand{\allUipEx}{\textit{alluip-exclusive}}
\newcommand{\MapleBase}{\textit{MapleCOMSPS\_LRB}}
\newcommand{\MapleSeven}{\textit{MapleLCMDist}}
\newcommand{\MapleNine}{\textit{MapleLCMDiscChronoBT-DL-v3} }
\newcommand{\MapleEight}{\textit{MapleLCMDiscChronoBT}}
\newcommand{\MapleNineShort}{\textit{MapleCB-DL} }
\newcommand{\expSAT}{\textit{expMaple-CM-GCBumpOnlyLRB} }
\newcommand{\expSATShort}{\textit{expMaple} }
\newcommand{\MapleIUIPPure}{\text{Maple-\allUipPure}}
\newcommand{\MapleIUIMin}{\text{Maple-\allUipMin}}
\newcommand{\MapleEightShort}{\textit{MapleCB}}
\newcommand{\Set}[2]{\{\,#1\mid#2\,\}}
\newcommand{\Array}[2]{[\,#1\mid#2\,]}
\newcommand{\trail}{\ensuremath{\mathcal{T}}}
\newcommand{\trailIdx}[1]{\ensuremath{\iota(#1)}}
\newcommand{\dlevel}[1]{\ensuremath{\mathit{decLvl}(#1)}}
\newcommand{\dlevels}{\ensuremath{\mathit{decLvls}}}
\newcommand{\var}{\text{var}}
\newcommand{\true}{\textsc{true}\xspace}
\newcommand{\false}{\textsc{false}\xspace}
\newcommand{\reason}[1]{\ensuremath{\mathit{reason}(#1)}}
\newcommand{\resolve}{\bowtie}
\renewcommand{\implies}{\rightarrow}
\newcommand{\ctry}{C_{\mathit{try}}}
\newcommand{\type}{\textit{type}}
\newcommand{\unmarked}{\textbf{unmarked}}
\newcommand{\mrk}{\textbf{mark}}
\setlength{\theorempreskipamount}{3pt}
\setlength{\theorempostskipamount}{3pt}

@@

\title{Clause Size Reduction with \alluip Learning}
\author{Nick Feng \and Fahiem Bacchus}
\authorrunning{N. Feng and F. Bacchus}
\institute{Department of Computer Science, University of Toronto, Canada, 
  \email{\{fengnick,fbacchus\}@cs.toronto.edu}}

\begin{document}
\maketitle              % typeset the header of the contribution
% 
\begin{abstract}
    Almost all CDCL \sat solvers use the \oneuip 
    clause learning scheme for learning new clauses from conflicts, 
    and our current understanding
    of \sat solving provides good reasons for using that scheme. In
    particular, the \oneuip scheme yields asserting clauses, and these
    asserting clauses have minimum \LBD among all possible asserting
    clauses. As a result of these advantages, other clause learning
    schemes, like $i$-UIP and \alluip, that were proposed in early
    work are not used in modern solvers. In this paper, we propose a
    new technique for exploiting the \alluip clause learning scheme. 
    Our technique is to employ \alluip learning under the
    constraint that the learnt clause's \LBD does not increase (over the
    minimum established by the \oneuip clause). Our method can learn
    clauses that are significantly smaller than the \oneuip clause while
    preserving the minimum \LBD. Unlike previous clause minimization
    methods, our technique is not limited to learning a sub-clause of
    the \oneuip clause. It can instead learn a completely different
    clause. We show empirically 
    that our method can improve the performance of state of the art solvers.
\end{abstract}

\section{Introduction}
Clause learning is an essential technique in \sat solvers. There is
good evidence to indicate that it is, in fact, the most important
technique used in modern \sat solvers
\cite{DBLP:conf/sat/KatebiSS11}. In early \sat research a number of
different clause learning techniques were proposed
\cite{DBLP:conf/iccad/ZhangMMM01,DBLP:conf/iccad/SilvaS96,DBLP:journals/tc/Marques-SilvaS99,DBLP:conf/aaai/BayardoS97}.
However, following the revolutionary performance improvements achieved
by the Chaff \sat solver, the field has converged on using the \oneuip
(first Unique Implication Point) scheme
\cite{DBLP:conf/iccad/ZhangMMM01} employed in Chaff
\cite{DBLP:conf/dac/MoskewiczMZZM01} (as well as other techniques
pioneered in the Chaff solver).\footnote{The idea of UIP clauses was
  first mentioned in \cite{DBLP:conf/iccad/SilvaS96}, and
  \oneuip clauses along with other UIP clauses were learnt and used in
  the earlier GRASP \sat solver.} Since then almost all \sat solvers
have employed the \oneuip clause learning scheme, along with clause
minimization \cite{DBLP:conf/sat/SorenssonB09}, as their primary
method for learning new clauses.

However, other clause learning schemes can be used in SAT solvers
without changes to the main data structures. Furthermore, advances in
our understanding allow us to better understand the potential
advantages and disadvantages of these alternate schemes. In this paper
we reexamine these previously proposed schemes with a focus on the
schemes described in \cite{DBLP:conf/iccad/ZhangMMM01}. Improved
understanding of \sat solvers, obtained from the last decade of
research, allows us to see that in their original form these
alternative clause learning schemes suffer significant disadvantages
over \oneuip clause learning.

One of the previously proposed schemes was the \alluip scheme
\cite{DBLP:conf/iccad/ZhangMMM01}. In this paper we propose a new way
to exploit the main ideas of this scheme that avoids its main
disadvantage which is that it can learn clauses with higher \LBD
scores. In particular, we propose to use a \alluip like clause
learning scheme to generate smaller learnt clauses which retain the
good properties of standard \oneuip clauses. Our method is related to,
but not the same as, various clause mininimization methods that try to
remove redundant literals from the \oneuip clause yielding a clause that
is a subset of the \oneuip clause, e.g.,
\cite{DBLP:conf/sat/SorenssonB09,DBLP:conf/ijcai/LuoLXML17,DBLP:conf/sat/WieringaH13}.
Our method is orthogonal to clause minimization. In particular, our
approach can learn a clause that is not a subset of the \oneuip clause
but which still serves all of the same purposes as the \oneuip clause.
Various minimization techniques can be applied on top of our method to
remove redundant literals from the clauses we have learnt.

We present various versions of our method and show that these variants
are often capable of learning shorter clauses than the \oneuip scheme,
and that this can lead to useful performance gains in state of the art
\sat solvers.

@@

Two types of true literals appear on the trail: \emph{decision
  literals} that have been assumed to be true by the solver, and
\emph{unit propagated literals} that are \emph{forced to be true}
because they are the sole remaining unfalsified literal of a
clause. Each literal $\ell\in\trail$ has a decision level
$\dlevel{\ell}$. Let $k$ be the number of decision literals appearing
before $\ell$ on the trail. When $\ell$ is a unit propagated literal
$\dlevel{\ell}=k$, and when $\ell$ is a decision literal
$\dlevel{\ell}=k+1$. For example, $\dlevel{d}=1$ for the first
decision literal $d\in\trail$, and $\dlevel{\ell}=0$ for all literals
$\ell$ appearing before $d$ on the trail.  The set of literals on
$\trail$ that have the same decision level forms a contiguous
subsequence of $\trail$ that starts with a decision literal $d_i$ and
ends just before the next decision literal $d_{i+1}$. If
$\dlevel{d_i} = i$ we call this subsequence of $\trail$ the
\textit{$i$-th decision level}.

@@

In most \sat solvers, clause learning is initiated as soon as a clause
is falsified by $\trail$. In this paper we will be concerned with the
subsequent clause learning process which uses $\trail$ to derive a new
clause. We will try to make as few assumptions about how $\trail$ is
managed by the \sat solver as possible. One assumption we will make is
that $\trail$ \emph{remains intact during clause learning} and is only
changed after the new clause is learnt. After the new clause is learnt
$\trail$ will be changed by backtracking.

Say that $\trail$ falsifies a clause $C_I$, and that the last decision
literal $d_k$ in $\trail$ has decision level $k$. Consider
$\trail_{k-1}$ the prefix of $\trail$ above the last decision level,
i.e., the sequence of literals
$\trail[0]$---$\trail[\trailIdx{d_k}-1]$. We will assume that
$\trail_{k-1}$ is \textbf{unit propagation complete}, although the
full trail $\trail$ might not be. This means that (a) no clause was
falsified by $\trail_{k-1}$. And (b) if $C_u$ is a clause containing
the literal $x$ and all literals in $C_u$ except for $x$ are falsified
by $\trail_{k-1}$, then $x\in \trail_{k-1}$ and
$\dlevel{x} \leq\max\{\dlevel{y} | y\in C_u \land y\neq x\}$. This
means that if $x$ appears in a clause made unit it must have been
added to the trail, and added at or before decision level the clause
became unit. Note that more than one clause might be made unit by
$\trail$ forcing $x$, or $x$ might be set as a decision before being
forced. This condition ensures that $x$ appears in $\trail$ at or
before the first decision level it is forced by any clause.

Any clause falsified by $\trail$ is called a \textbf{conflict}. When a
conflict is found, the final level of the trail, $k$, need not be unit
propagation complete as the solver typically stops propagation as soon
as it finds a conflict. This means that (a) other clauses might be
falsified by $\trail$ besides the conflict found, and (b) other
literals might be unit implied by $\trail$ but not added to $\trail$.

@@

\begin{Obs}
    \label{obs:ordered}
    If the unordered trail resolution $U$ yields the conflict clause
    $C_L$ from an initial conflict $C_I$, then there exists an ordered
    trail resolution $O$ that yields a conflict clause $C'_L$ such
    that $C'_L\subseteq C_L$.
\end{Obs}
\noindent
\textbf{Proof.} Let $U$ be the sequence of clauses $C_I = C_0$, $C_1$,
$\ldots$, $C_m= C_L$ obtained by resolving on the sequence of
variables $v_1$, $\ldots$, $v_m$ whose corresponding literals on
$\trail$ are $l_1$, $\ldots$, $l_m$. Reordering these resolution steps
so that the variables are resolved in order of decreasing trail index
and removing duplicates yields an ordered trail resolution $O$ with
the desired properties. Since no reason clause contains literals with
higher trail indices, $O$ must be a valid trail resolution if $U$
was, and furthermore $O$ yields the clause
$C'_L = \bigcup_{i=1}^m \reason{l_i} - \{l_1, \lnot l_1, \ldots, l_m,
\lnot l_m\}$. Since $U$ resolves on the same variables (in a different
order) using the same reason clauses we must have $C'_L\subseteq
C_L$. It can, however, be the case that $C'_L$ is proper subset of
$C_L$: if $l_i$ is resolved away it might be reintroduced when
resolving on $l_{i+1}$ if $\trailIdx{l_{i+1}}  > \trailIdx{l_i}$.
\whitebox

The relevance of trail resolutions is that all proposed clause
learning schemes we are aware of use trail resolutions to produce
learnt clauses. Furthermore, the commonly used technique for clause
minimization \cite{DBLP:conf/sat/SorenssonB09} is also equivalent to a
trail resolution that yields the minimized clause from the
un-minimized clause. Interestingly, it is standard in \sat solver
implementations to perform resolution going backwards along the
trail. That is, these implementations are typically using ordered
trail resolutions. Observation~\ref{obs:ordered} shows that this in
fact the right way to do this.

@@

\subsection{Some Alternate Clause Learning Schemes}
A number of different clause learning schemes for generating a new
learnt clause from the initial conflict have been presented in prior
work, e.g.,
\cite{DBLP:conf/iccad/ZhangMMM01,DBLP:conf/iccad/SilvaS96,DBLP:journals/tc/Marques-SilvaS99,DBLP:conf/aaai/BayardoS97}.
Figure~\ref{fig:cl_schemes} gives a specification of some of these
methods: (a) the all-decision scheme which resolves away all implied
literals leaving a learnt clause over only decision literals; (c) the
\oneuip scheme which resolves away literals from the deepest decision
level leaving a learnt clause with a single literal at the deepest
level; (d) the \alluip scheme which resolves away literals from each
decision level leaving a learnt clause with a single literal at each
decision level; and (e) the $i$-UIP scheme which resolves away
literals from the $i$ deepest decision levels leaving a learnt clause
with a single literal at its $i$ deepest decision levels. It should be
noted that when resolving away literals at decision level $i$, new
literals at decision levels less than $i$ might be introduced into the
clause. Hence, it is important in the $i$-UIP and \alluip schemes to
use ordered trail resolutions.

Both the all-decision and \alluip schemes yield a clause with only one
literal at each decision level, and the \alluip clause will be no
larger that the all-decision clause. Furthermore, it is known
\cite{DBLP:journals/tc/Marques-SilvaS99} that once we reduce the
number of literals at a decision level $d$ to one, we could continue
performing resolutions and later achieve a different single literal at
the level $d$. In particular, a decision level might contain more than
one unique implication point, and in some contexts the term all-UIP is
used to refer to all the unique implication points that exist in a
particular decision level \cite{DBLP:conf/sat/SabharwalSS12} rather
than the \alluip clause learning scheme as is used here. The
algorithms given in Figure~\ref{fig:cl_schemes} stop at the first UIP
of a level, except for the all-decision schemes with stops at the last
UIP of each level.

\begin{figure}[t!]
{\fss{9pt}{10pt}
\begin{tabular}[t]{|l|c|l|}
  \cline{1-1}\cline{3-3}
  \begin{minipage}[t]{0.48\textwidth}
    \textbf{(a) All Decision Clause\rule{0pt}{1.1\topskip}}
    \begin{algorithmic}[0]
        \State \hspace*{-1em}\textbf{all\_decision}($C_I$)
        \State $C\gets C_I$
        \While{$\{l\,|\, l\in C \land\reason{l} \neq\varnothing\} \neq \emptyset$}
        \State \begin{tabular}{ll}
                 $\ell \gets \mbox{}$ & literal with highest trail index in\\
                                      & $\{l\,|\, l\in C \land\reason{l} \neq\varnothing\}$
               \end{tabular}
        \State $C \gets C\resolve \reason{\lnot \ell}$
        \EndWhile
        \State \textbf{return} $C$
    \end{algorithmic}
  \end{minipage}
  & &
  \begin{minipage}[t]{0.475\textwidth}
      \textbf{(b) Make level $i$ contain a single literal\rule{0pt}{1.1\topskip}}
      \begin{algorithmic}[0]
          \State \hspace*{-1em} \textbf{UIP\_level}($C$, $i$)
          \While{$\left|\left\{\ell \left| \begin{array}{l}\ell \in C\\
                                           \mbox{}\land \reason{l} \neq \varnothing \\
                                           \mbox{}\land \dlevel{l} = i\end{array}\right.\right\}\right| > 1$}
          \State \begin{tabular}{ll}
                   $l \gets \mbox{}$ & literal with highest trail index in\\
                                        & $\left\{\ell \left| \begin{array}{l}\ell \in C
                                          \land \reason{l} \neq \varnothing \\
                                          \mbox{}\land \dlevel{l} = i\end{array}\right.\right\}$
                 \end{tabular}
          \State $C \gets C\resolve \reason{\lnot l}$
          \EndWhile
          \State \textbf{return} $C$\strut
      \end{algorithmic}
  \end{minipage}
  \\\cline{1-1}\cline{3-3}
   \multicolumn{1}{c}{\vspace*{-7pt}}\\\cline{1-1}\cline{3-3}
  \begin{minipage}[t]{0.475\textwidth}
    \begin{minipage}{\textwidth}
        \textbf{(c) First UIP Clause \rule{0pt}{1.1\topskip}}
        \begin{algorithmic}[0]
            \State \hspace*{-1em}\textbf{\oneuip}($C_I$)
            \State $i\gets \max\{\dlevel{l}\, |\, l \in C_I\}$
            \State \textbf{return} \textbf{UIP\_level}($C_I$, i)
        \end{algorithmic}
        \vspace*{5pt}
    \end{minipage}\\[5pt]
    \hspace*{-1.8pt}\rule{1.03\textwidth}{.4pt}\\[5pt]
    \begin{minipage}{\textwidth}
        \textbf{(d) \alluip Clause \rule{0pt}{1.1\topskip}}
        \begin{algorithmic}[0]
          \State \hspace*{-1em}\textbf{\alluip}($C_I$, $i$)
          \State $i = \bigl|\{\dlevel{l} \,|\, l \in \trail\}\bigl|$
          \State \Comment{$i$ is large enough to ensure all levels are UIP}
          \State \textbf{return} \textbf{$i$-UIP}(C, i)

@@

\subsection{Asserting Clauses and \LBD---Reasons to Prefer \oneuip
  Clauses}
An \textbf{asserting clause} \cite{DBLP:journals/ai/PipatsrisawatD11}
is a conflict clause $C_L$ that has exactly one literal $\ell$ at its
deepest level, i.e.,
$\forall x\in C_L. \dlevel{x} \leq \dlevel{\ell} \land (\dlevel{x} =
\dlevel{\ell} \implies x = \ell$). All of the clause learning schemes in
Figure~\ref{fig:cl_schemes} produced learnt clauses that are asserting.

The main advantage of asserting clauses is that they are 1-Empowering
\cite{DBLP:journals/ai/PipatsrisawatD11}, i.e., they allow unit
propagation to derive a new forced literal. Hence, asserting clauses
can be used to guide backtracking---the solver can backtrack from the
current deepest level to the point the learnt clause first becomes
unit, and then use the learnt clause to add a new unit implicant to
the trail. Since all but the deepest level was unit propagation
complete, this means that the asserting clause must be a brand new
clause; otherwise that unit implication would already have been
made. On the other hand, if the learnt clause $C_L$ is not asserting
then it could be that it is a duplicate of another clause already in
the formula.
